---
title: "Lab 09: Algorithmic Bias"
author: "Sophie Boyd"
date: "2-27-26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r read-data, message=FALSE}
compas <- read_csv("data/compas-scores-2-years.csv")
```

```{r clean-data}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```


### Exercise 1

```{r dimensions}
nrow(compas)
ncol(compas)
```

The datset contains 7214 observations of 53 variables. Each row represents a defendant and the variables provide demographic information about the defendants and details about their charges, criminal history,  estimated future risk of recidivism, and actual rates of recidivism.

### Exercise 2

As far as I know, each row represents a unique defendant. I might be missing something...

### Exercise 3

```{r dist-risk-scores}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.5) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count')
```
The distribution is left-skewed. The majority of the defendants had a low risk score, and the numbers of defendants with higher scores follow a downward trend.

### Exercise 4

```{r dist-risk-race}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~race, nrow = 6) 
```

```{r dist-risk-sex}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~sex, nrow = 2) 
```

```{r dist-risk-age-cat}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~age_cat, nrow = 6) 
```

### Exercise 5

```{r risk-recid}
compas %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~two_year_recid, nrow=2) +
  labs(x = 'COMPAS score',
       y = 'Count',
       )
```

Higher risk scores do not correspond to higher rates of recidivism. Among defendants who did recidivate, there are no meaningful differences in their COMPAS scores.

### Exercise 6

```{r classification}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 7 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 7 & two_year_recid == 1 ~ "FN"
  ))
```

### Exercise 7

```{r accuracy}
compas %>%
  count(compas_classification)
```

### Exercise 7

(3561 + 1351)/7214

The COMPAS score has a 68% accuracy rate.

### Exercise 8

```{r risk-black-white}
compas%>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~race, nrow=2) +
  labs(x = 'COMPAS score',
       y = 'Count',
       )
```

Compared to white defendants, Black defendants are vastly overrepresented in the high-risk scores.

### Exercise 9

```{r high-risk}
compas <- compas %>%
  mutate(highrisk = case_when(
    decile_score >= 7 ~ 1,
    decile_score < 7 ~ 0
  ))

compas %>%
  group_by(race) %>%
  filter(highrisk == 1) %>%
  count(highrisk) %>%
  ungroup()
```

```{r}
1425 / 1995

419/ 1995
```

Yes there is a disparity. Black defendants make up a much higher percentage of high-risk scores than white defendants. 

```{r}
non_recidivists <- compas %>%
  filter(two_year_recid == 0)
  
non_recidivists %>%
  count(highrisk)
```

```{r}
644 / (3319+644)
```

```{r}
non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  count(highrisk) %>%
  ungroup()
```

```{r}
compas <- compas %>%
  mutate(lowrisk = case_when(
    decile_score <=4 ~ 1,
    decile_score > 4 ~ 0
  ))

recidivists <- compas %>%
  filter(two_year_recid == 1)
  
recidivists %>%
  count(lowrisk)

recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  count(lowrisk) %>%
  ungroup()
```

```{r}
non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = highrisk, group = race)) +
  geom_bar() +
  facet_wrap(~race, nrow=2)
```


