---
title: "Lab 09: Algorithmic Bias"
author: "Sophie Boyd"
date: "2-27-26"
output: github_document
---

## search I did for exercise 12: 
https://www.google.com/search?q=how+to+restrict+range+of+x+values+in+ggplot2&sca_esv=11f77b5468a41098&rlz=1C5CHFA_enUS912US912&sxsrf=ANbL-n4n0DKV07oSujm7po9TBUMRugrUTA%3A1772208420436&ei=JMGhaarjGb6K6-AP2tTtkQ0&biw=1440&bih=716&oq=how+to+restrict+range+of+x+values+&gs_lp=Egxnd3Mtd2l6LXNlcnAiImhvdyB0byByZXN0cmljdCByYW5nZSBvZiB4IHZhbHVlcyAqAggDMgUQIRigATIFECEYoAEyBRAhGKABMgUQIRigATIFECEYoAEyBRAhGJ8FMgUQIRifBTIFECEYnwUyBRAhGJ8FMgUQIRifBUiYU1DsBljlQnAFeAGQAQCYAZ8BoAHfG6oBBTE5LjE3uAEByAEA-AEBmAIpoAK5HsICChAAGLADGNYEGEfCAgsQABiABBiRAhiKBcICDhAuGIAEGLEDGNEDGMcBwgIKEAAYgAQYQxiKBcICCxAuGIAEGNEDGMcBwgIOEC4YgAQYsQMYgwEYigXCAggQABiABBixA8ICChAuGIAEGEMYigXCAg0QLhiABBhDGOUEGIoFwgIOEAAYgAQYsQMYgwEYigXCAgUQABiABMICBRAuGIAEwgILEAAYgAQYsQMYigXCAggQLhiABBixA8ICCxAAGIAEGLEDGIMBwgIGEAAYFhgewgIFECEYqwLCAgcQIRigARgKmAMAiAYBkAYIkgcFMjAuMjGgB8TJArIHBTE1LjIxuAeQHsIHBjItMzYuNcgHhgKACAA&sclient=gws-wiz-serp

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r read-and-clean-data, message=FALSE}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )
```


### Exercise 1

```{r dimensions}
nrow(compas)
ncol(compas)
```

The datset contains 7214 observations of 53 variables. Each row represents a defendant. The variables provide demographic information about the defendants and details about their charges, criminal history, estimated future risk of recidivism, and actual rates of recidivism.

### Exercise 2

As far as I can tell, each row represents a unique defendant. (I could be missing something!)

### Exercise 3

```{r dist-risk-scores}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.5) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count')
```
The distribution is right-skewed. The majority of the defendants had a low risk score, and the numbers of defendants with higher scores follow a downward trend.

### Exercise 4

```{r dist-risk-race}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~race, nrow = 6) 
```

```{r dist-risk-sex}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~sex, nrow = 2) 
```

```{r dist-risk-age-cat}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_histogram(binwidth=.4) + 
  labs(x = 'COMPAS risk scores',
       y = 'Count') +
  facet_wrap(~age_cat, nrow = 6) 
```

### Exercise 5

```{r risk-recid}
compas %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~two_year_recid, nrow=2) +
  labs(x = 'COMPAS score',
       y = 'Count',
       )
```

Higher risk scores do not correspond to higher rates of recidivism. Among defendants who did recidivate (the lower plot), there are no meaningful differences in their COMPAS scores.

### Exercise 6

```{r classification}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 7 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 7 & two_year_recid == 1 ~ "FN"
  ))
```

### Exercise 7

```{r accuracy}
compas %>%
  group_by(compas_classification) %>%
  summarise(n = n()) %>%
  mutate(frequency = n / sum(n)) %>%
  ungroup()
```

The COMPAS score has a 68% accuracy rate (combined frequencies of true positives and true negatives).

### Exercise 8

```{r risk-black-white}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~race, nrow=2) +
  labs(x = 'COMPAS risk score',
       y = 'Count',
       )
```

Compared to white defendants, Black defendants are vastly overrepresented in the high-risk scores.

### Exercise 9

```{r high-risk}
compas <- compas %>%
  mutate(highrisk = case_when(
    decile_score >= 7 ~ 1,
    decile_score < 7 ~ 0
  ))

compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, highrisk) %>%
  summarise(n = n()) %>%
  mutate(frequency = n / sum(n)) %>%
  ungroup()
```

Yes there is a disparity. 38.6% of Black defendants were classified as high risk, whereas only 17.1% of white participants were classified as high risk. 

### Exercise 10

```{r FP-rates}
non_recidivists <- compas %>%
  filter(two_year_recid == 0) 

non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))
```

```{r FN-rates}
recidivists <- compas %>%
  filter(two_year_recid == 1) 

recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))
```

### Exercise 11

```{r FP-plot-by-race}
non_recidivists%>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = compas_classification, y = prop, color = compas_classification, fill = compas_classification)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~race, nrow=2)
```

```{r FN-plot-by-race}
recidivists%>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = compas_classification, y = prop, color = compas_classification, fill = compas_classification)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~race, nrow=2)
```

Compared to white defendants, Black defendants had a higher false positive rate and a lower false negative rate, meaning that their recidivism risk was more likely to be overestimated and less likely to be underestimated.

### Exercise 12

```{r frequ-priors}
compas %>%
  count(priors_count, desc. = TRUE)

compas <- compas %>%
  mutate(cont_decile_score = as.numeric(as.character(decile_score)))
```


```{r risk-by-priors}
compas %>%
 ggplot(aes(x = priors_count, y = cont_decile_score, color = race, fill = race)) +
  geom_point() 
```

### Exercise 13

```{r calibration}
compas %>%
  group_by(race, decile_score) %>%
  summarise()
  mutate(recid_rate = )
```

