---
title: "Lab 09: Algorithmic Bias"
author: "Sophie Boyd"
date: "2-27-26"
output: github_document
---

## search I did for exercise 12: 
https://www.google.com/search?q=how+to+restrict+range+of+x+values+in+ggplot2&sca_esv=11f77b5468a41098&rlz=1C5CHFA_enUS912US912&sxsrf=ANbL-n4n0DKV07oSujm7po9TBUMRugrUTA%3A1772208420436&ei=JMGhaarjGb6K6-AP2tTtkQ0&biw=1440&bih=716&oq=how+to+restrict+range+of+x+values+&gs_lp=Egxnd3Mtd2l6LXNlcnAiImhvdyB0byByZXN0cmljdCByYW5nZSBvZiB4IHZhbHVlcyAqAggDMgUQIRigATIFECEYoAEyBRAhGKABMgUQIRigATIFECEYoAEyBRAhGJ8FMgUQIRifBTIFECEYnwUyBRAhGJ8FMgUQIRifBUiYU1DsBljlQnAFeAGQAQCYAZ8BoAHfG6oBBTE5LjE3uAEByAEA-AEBmAIpoAK5HsICChAAGLADGNYEGEfCAgsQABiABBiRAhiKBcICDhAuGIAEGLEDGNEDGMcBwgIKEAAYgAQYQxiKBcICCxAuGIAEGNEDGMcBwgIOEC4YgAQYsQMYgwEYigXCAggQABiABBixA8ICChAuGIAEGEMYigXCAg0QLhiABBhDGOUEGIoFwgIOEAAYgAQYsQMYgwEYigXCAgUQABiABMICBRAuGIAEwgILEAAYgAQYsQMYigXCAggQLhiABBixA8ICCxAAGIAEGLEDGIMBwgIGEAAYFhgewgIFECEYqwLCAgcQIRigARgKmAMAiAYBkAYIkgcFMjAuMjGgB8TJArIHBTE1LjIxuAeQHsIHBjItMzYuNcgHhgKACAA&sclient=gws-wiz-serp

## exercise 3:
https://ggplot2.tidyverse.org/articles/faq-axes.html#:~:text=Dodge%20axis%20labels:%20Add%20a,rows%20to%20render%20the%20labels.

## exercise 7:

proportions_df <- df %>%
  group_by(category) %>%
  summarize(count = n()) %>%
  mutate(proportion = count / sum(count)) #

print(proportions_df)

https://datavizm20.classes.andrewheiss.com/lesson/04-lesson/#:~:text=When%20you%20visualize%20proportions%20with%20ggplot%2C%20you'll,summarize()%20)%20*%20Plot%20the%20summarized%20data.
https://collinn.github.io/f24/labs/04_dplyr.html#:~:text=Whenever%20functions%20from%20the%20dplyr%20package%20see,it%20intends%20to%20at%20the%20group%20level.
https://www.r-bloggers.com/2024/02/group-by-minimum-in-r/#:~:text=It%20(%20The%20GROUP%20BY%20clause%20),to%20perform%20these%20operations%20on%20data%20frames.
scaler.com/topics/dplyr-package-in-r/#:~:text=As%20R%20users%20commonly%20work%20with%20data,a%20more%20natural%20workflow%2C%20simplifying%20data-wrangling%20tasks.

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r read-and-clean-data, message=FALSE}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )
```


### Exercise 1

```{r dimensions}
nrow(compas)
ncol(compas)
```

The datset contains 7214 observations of 53 variables. Each row represents a defendant. The variables provide demographic information about the defendants and details about their charges, criminal history, estimated future risk of recidivism, and actual rates of recidivism.

### Exercise 2

As far as I can tell, each row represents a unique defendant. (I could be missing something!)

### Exercise 3

```{r dist-risk-scores}
compas %>%
  ggplot(aes(x = decile_score)) +
  geom_bar() + 
  labs(x = 'COMPAS risk scores',
       y = 'Count',
       title = 'Distribution of COMPAS Risk Scores') +
    scale_x_continuous(breaks = seq(1,10, by =1), labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
```
The distribution is right-skewed. The majority of the defendants had a low risk score, and the numbers of defendants with higher scores follow a downward trend.

### Exercise 4

```{r dist-race}
compas <- compas %>%
  mutate(race_ordered = fct_relevel(race, 'African-American', 'Caucasian', 'Hispanic', 'Other', 'Asian', 'Native American'))

compas %>%
  ggplot(aes(x = race_ordered)) +
  geom_bar() + 
  labs(x = 'Race',
       y = 'Count',
       title = 'Distribution of Defendants by Race') 
```

```{r dist-sex}
compas <- compas %>%
  mutate(sex_ordered = fct_relevel(sex, 'Male', 'Female'))
  
compas %>%
  ggplot(aes(x = sex_ordered)) +
  geom_bar() + 
  labs(x = 'Sex',
       y = 'Count',
       title = 'Distribution of Defendants by Sex') 
```

```{r dist-age-cat}
compas <- compas %>%
  mutate(age_cat_ordered = fct_relevel(age_cat, 'Less than 25', '25 - 45', 'Greater than 45'))

compas %>%
  ggplot(aes(x = age_cat_ordered)) +
  geom_bar() + 
  labs(x = 'Age Category',
       y = 'Count',
       title = 'Distribution of Defendants by Age Category') 
```

### Exercise 5

```{r risk-recid}
compas <- compas %>%
  mutate(recid_label = case_when(
    two_year_recid == 0 ~ "Did not recidivate",
    two_year_recid == 1 ~ "Recidivated"
  ))

compas %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~recid_label, nrow=2) +
  labs(x = 'COMPAS score',
       y = 'Count',
       ) +
  scale_x_continuous(breaks = seq(1,10, by =1), labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))
```

Higher risk scores do not correspond to higher rates of recidivism. Among defendants who did recidivate (the lower half of the plot), there are no meaningful differences in their COMPAS scores.

### Exercise 6

```{r classification}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN"
  ))
```

### Exercise 7

```{r accuracy}
compas %>%
  group_by(compas_classification) %>%
  summarise(n = n()) %>%
  mutate(frequency = n / sum(n)) %>%
  ungroup()
```

The COMPAS score has a 55% accuracy rate (combined frequencies of true positives and true negatives).

### Exercise 8

```{r risk-black-white}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = decile_score)) + 
  geom_bar() +
  facet_wrap(~race, nrow=2) +
  labs(x = 'COMPAS risk score',
       y = 'Count',
       )
```

Compared to white defendants, Black defendants are vastly overrepresented in the high-risk scores.

### Exercise 9

```{r high-risk}
compas <- compas %>%
  mutate(highrisk = case_when(
    decile_score >= 7 ~ 1,
    decile_score < 7 ~ 0
  ))

compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, highrisk) %>%
  summarise(n = n()) %>%
  mutate(frequency = n / sum(n)) %>%
  ungroup()
```

Yes there is a disparity. 38.6% of Black defendants were classified as high risk, whereas only 17.1% of white participants were classified as high risk. 

### Exercise 10

```{r FP-rates}
non_recidivists <- compas %>%
  filter(two_year_recid == 0) 

non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))

non_recidivists_prop <- non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))
```
The false positive rate among Black defendants was 24.9%, compared to a rate of just 9.1% among white defendants. 

```{r FN-rates}
recidivists <- compas %>%
  filter(two_year_recid == 1) 

recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))

recidivists_prop <- recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, compas_classification) %>%
  summarise(n=n()) %>%
  mutate(frequency = n / sum(n))
```
The false negative rate among white defendants (47.7%) was considerably higher than the false negative rate among Black defendants (28.0%).

### Exercise 11

```{r FP-plot-by-race}
non_recidivists_prop%>%
  filter(compas_classification %in% "FP") %>%
  ggplot(aes(x = race, y = frequency)) +
  labs(x = 'Race',
       y = 'False Positive Rate') +
  geom_col() 
```

```{r FN-plot-by-race}
recidivists_prop%>%
  filter(compas_classification %in% c("FN")) %>%
  ggplot(aes(x = race, y = frequency)) +
  labs(x = 'Race',
       y = 'False Negative Rate') +
  geom_col() 
```

Compared to white defendants, Black defendants had a higher false positive rate and a lower false negative rate, meaning that their recidivism risk was more likely to be overestimated and less likely to be underestimated.

### Exercise 12

```{r risk-by-priors, fig.height=10}
compas %>%
  ggplot(aes(x = priors_count, y = decile_score, color = race)) +
  geom_jitter() +
  facet_wrap(~race, nrow = 6)
```

Priors appear to be most influential on COMPAS scores for Black and Native American defendants.

### Exercise 13

```{r calibration}
compas$decile_cat <- as.factor(compas$decile_score)
  
compas%>%
  group_by(decile_cat, race, is_recid) %>%
  summarise(n = n()) %>%
  mutate(recid_rate = n / sum(n))
```

